\chapter{Introduction}

\begin{quote}
\textit{``The important thing is not to stop questioning. Curiosity has its own reason for existing.''} \\
\hspace*{\fill}--- Albert Einstein
\end{quote}

Curiosity is widely viewed as a driving force behind learning, both in humans and in artificial systems. In reinforcement learning (RL), this idea has been formalized as intrinsic motivation: an internally generated signal that encourages an agent to seek out novel or informative experiences when external feedback is scarce. Yet the question of \textit{how strongly} an agent should listen to its own curiosity remains largely unresolved. Most existing methods rely on a single, manually chosen coefficient that controls the influence of intrinsic rewards uniformly across all states. This thesis takes the position that the importance of curiosity is itself state-dependent and proposes a learned scaling mechanism that adapts the influence of intrinsic motivation to the current situation of the agent.

\section{Background and Research Motivation}
Reinforcement learning has emerged as a central paradigm for training agents to make sequential decisions in complex environments, with influential successes in Atari games, Go, and robotic control \cite{mnih2015human,silver2016go,sutton2018rl}. Many of these results, however, depend on dense reward signals that provide frequent feedback about which behaviors are desirable. In contrast, a wide range of practical environments are sparse-reward: the agent receives meaningful feedback only after a long sequence of actions, which makes exploration substantially harder \cite{francois2018deep}.

To address this difficulty, a large body of research augments the extrinsic reward supplied by the environment with an additional intrinsic reward that encourages exploration \cite{oudeyer2007intrinsic,schmidhuber2010formal}. Count-based methods reward the agent for visiting rarely seen states \cite{bellemare2016count,tang2017exploration,ostrovski2017count}. Curiosity-driven methods derive intrinsic rewards from prediction errors, representation discrepancies, or state-entropy estimates, as in the Intrinsic Curiosity Module (ICM) \cite{pathak2017curiosity}, Random Network Distillation (RND) \cite{burda2019rnd}, Rewarding Impact-Driven Exploration (RIDE) \cite{raileanu2020ride}, and Random Encoders for Efficient Exploration (RE3) \cite{seo2021re3}. These approaches improve exploration in many tasks, but they all share a key design choice: the strength with which intrinsic rewards influence learning relative to extrinsic rewards.

In standard practice, this strength is controlled by a fixed scalar coefficient that must be tuned manually for each environment. The optimal value of this coefficient often varies across tasks and even across different stages of training \cite{chen2022eipo,yuan2023airs}. More fundamentally, a single global coefficient cannot distinguish between states where exploration remains useful and states where exploration has become distracting or harmful. The same limitation appears across multiple intrinsic reward families, from count-based bonuses to ICM and RIDE. This shared weakness motivates a broader research perspective: rather than studying yet another curiosity module, this thesis investigates an adaptive mechanism whose effect can be analyzed across representative intrinsic reward modules.

\section{Research Problem}
This thesis investigates the following problem: how does adaptive, state-dependent intrinsic reward scaling affect different intrinsic reward modules, and can such scaling promote useful exploration in sparse-reward reinforcement learning without destabilizing policy optimization?

More specifically, let the agent receive an extrinsic reward $R_t^E$ from the environment and an intrinsic reward $R_t^{I,m}$ produced by intrinsic reward module $m$, where $m$ may denote a count-based bonus, ICM, or RIDE. Standard methods optimize a shaped reward of the form
\begin{equation}
\bar{r}_t = R_t^E + \beta R_t^{I,m},
\end{equation}
where $\beta$ is a fixed hyperparameter. The limitation of this formulation is that the same intrinsic scaling is applied uniformly to all states, even though the value of exploration may differ substantially across the state space. The central research question of this thesis is therefore:
\begin{quote}
\textit{Can a state-dependent scaling mechanism, learned online from the agent's own experience, improve how representative intrinsic reward modules such as count-based bonuses, ICM, and RIDE contribute to learning in sparse-reward environments?}
\end{quote}

\section{Research Objectives and Scope}
The objective of this thesis is to develop and empirically evaluate \textbf{Correlation-Aware Reward Exploration (CARE)}, a state-dependent scaling mechanism for intrinsic rewards, and to study its effect on representative intrinsic reward modules in sparse-reward reinforcement learning. The specific objectives are as follows:
\begin{itemize}
    \item analyze why fixed intrinsic reward coefficients are a shared limitation across representative intrinsic reward modules;
    \item review count-based exploration, ICM, RIDE, and prior adaptive reward weighting approaches as the foundation for state-dependent adaptation;
    \item formulate CARE as a lightweight, module-agnostic scaling layer that learns a state-dependent coefficient $\beta(s_t)$ via a correlation objective that couples the weighted intrinsic reward to the extrinsic GAE advantage of the value function;
    \item describe the concrete implementation of CARE with Proximal Policy Optimization (PPO) and three representative intrinsic reward modules (count-based bonuses, ICM, and RIDE), each compared against a sweep of fixed scalar coefficients;
    \item evaluate the empirical effect of adaptive scaling in sparse-reward MiniGrid environments and discuss what the results imply for different intrinsic reward modules.
\end{itemize}

The scope of this thesis is limited to episodic reinforcement learning tasks with sparse rewards. Conceptually and empirically, the study is framed around three representative intrinsic reward families: count-based bonuses, ICM, and RIDE. The quantitative experiments use PPO as the base policy optimization algorithm and instantiate the framework with all three intrinsic reward modules; for each module, the adaptive CARE variant is compared against a sweep of fixed scalar coefficients $\beta \in \{0.0005,\, 0.001,\, 0.005,\, 0.01,\, 0.05\}$ together with a no-intrinsic PPO control. The experiments focus on benchmark MiniGrid environments \cite{chevalier2018minigrid} rather than real-world robotics or continuous-control tasks. The thesis emphasizes empirical performance and methodological clarity rather than a full theoretical proof of convergence or optimality.

\section{Research Methodology}
The work follows a research-oriented methodology with four stages. First, the thesis surveys foundational RL literature and representative intrinsic reward modules to identify the common limitation of fixed intrinsic weighting. Second, it formulates CARE as an adaptive scaling mechanism that predicts a positive coefficient $\beta(s_t)$ from the current state and can be applied to module-specific intrinsic rewards. Third, the method is instantiated within a PPO training loop with three representative intrinsic reward modules (count-based bonuses, ICM, and RIDE). Finally, the approach is validated experimentally on multiple sparse-reward tasks in MiniGrid by comparing it against PPO without intrinsic rewards and against each module with a swept set of fixed intrinsic coefficients.

The evaluation criteria emphasize sample efficiency, learning stability, robustness across environments, and the qualitative behavior of the learned scaling function. This combination of theoretical grounding, algorithmic design, and empirical validation supports an experimental framing centered on the effect of CARE on intrinsic reward modules rather than on the construction of a new intrinsic reward generator.

\section{Thesis Structure}
The remainder of this thesis is organized as follows. Chapter 2 reviews reinforcement learning in sparse-reward environments, representative intrinsic reward modules, and prior attempts to adapt the weighting of intrinsic rewards; it also identifies the research gap that motivates the proposed method. Chapter 3 presents the proposed methodology, including the formal problem statement, the module-agnostic CARE formulation, the advantage-correlation training objective, the three intrinsic module instantiations (count-based, ICM, RIDE), and the complete training algorithm; it also states five formal results (Propositions, a Lemma, and a Corollary) that characterize CARE's properties. Chapter 4 describes the experimental setup, including the MiniGrid benchmark environments, the implementation details, the evaluation metrics, and the analysis of results from the perspective of CARE's effect on intrinsic reward modules. The conclusion summarizes the main findings, discusses limitations of the proposed approach, and outlines directions for future research. Two appendices complete the thesis: Appendix~A contains the full proofs of the formal results introduced in Chapter 3, and Appendix~B records the network architectures, hyperparameters, and training schedule used in the experiments.
