\chapter{Methodology}
\label{ch:methodology}

This chapter presents the proposed methodology in formal terms. It first defines the underlying intrinsic-augmented decision process and the overall system architecture. It then specifies the intrinsic reward generator used in the implemented system, formulates the Correlation-Aware Reward Exploration (CARE) framework with its Beta Network and learning-progress objective, and concludes with the complete training algorithm. Five formal results (Propositions 3.1--3.3, Lemma 3.1, Corollary 3.1) are stated in this chapter and proved in Appendix A.

\section{System Architecture}
\label{sec:system_architecture}

The proposed framework is built on top of an intrinsic-augmented Markov decision process (MDP) and introduces a learned, state-dependent scaling function for intrinsic rewards. The two ingredients are made explicit in the following definitions.

\begin{definition}[Intrinsic-Augmented MDP]
\label{def:augmented_mdp}
An \emph{intrinsic-augmented MDP} is a tuple $(\mathcal{S}, \mathcal{A}, P, R^E, R^I, \gamma, \rho_0)$ in which $(\mathcal{S}, \mathcal{A}, P, R^E, \gamma, \rho_0)$ defines a standard MDP with extrinsic reward function $R^E : \mathcal{S} \times \mathcal{A} \to \mathbb{R}$, and $R^I : \mathcal{S} \times \mathcal{A} \times \mathcal{S} \to \mathbb{R}_{\ge 0}$ is an additional, internally generated intrinsic reward function. The agent observes only $(s_t, a_t, R_t^E, s_{t+1})$ from the environment; the intrinsic reward $R_t^I$ is computed by the agent itself.
\end{definition}

\begin{definition}[State-Dependent Reward Shaping]
\label{def:reward_shaping}
A \emph{state-dependent reward shaping function} is a measurable map $\beta : \mathcal{S} \to \mathbb{R}_+$. Given an intrinsic-augmented MDP and a global intrinsic-strength hyperparameter $\alpha > 0$, the corresponding shaped reward at time step $t$ is
\begin{equation}
\bar{r}_t = R_t^E + \alpha\, \beta(s_t)\, R_t^I.
\label{eq:shaped_reward}
\end{equation}
The standard fixed-coefficient formulation corresponds to the special case $\beta(s) \equiv 1$.
\end{definition}

The proposed system instantiates this formulation with a parameterized $\beta_\psi$ together with one of three intrinsic reward generators --- count-based exploration, the Intrinsic Curiosity Module, or Random Network Distillation --- all of which are evaluated in Chapter~\ref{chap:experiment}. The full pipeline is shown in Figure~\ref{fig:care_arch}.

\begin{figure}[H]
\centering
\begin{tikzpicture}[
    >=latex,
    node distance=1.4cm,
    box/.style={draw, rounded corners, minimum width=2.5cm, minimum height=1.0cm, align=center, font=\small},
    op/.style={draw, circle, minimum size=0.7cm, font=\small, inner sep=1pt},
    arr/.style={->, thick},
    darr/.style={->, thick, dashed},
]
    % Top row: pipeline producing the intrinsic reward
    \node[box, fill=green!15] (env) {Environment};
    \node[box, fill=gray!15, right=1.6cm of env] (buf) {Rollout\\Buffer};
    \node[box, fill=blue!15, right=1.6cm of buf] (mod) {Intrinsic\\Module $m$};
    \draw[arr] (env) -- node[above, font=\footnotesize]{transitions} (buf);
    \draw[arr] (buf) -- node[above, font=\footnotesize]{$s_t,\,a_t,\,s_{t+1}$} (mod);

    % Middle row: Beta Network and shaped-reward composition
    \node[box, fill=orange!15, below=1.8cm of buf] (beta) {Beta Network\\$\beta_\psi(s_t)$};
    \node[op, right=1.0cm of beta] (mult) {$\times$};
    \node[op, right=1.0cm of mult] (plus) {$+$};
    \node[box, fill=red!10, right=1.2cm of plus] (ppo) {PPO Update};

    \draw[arr] (buf.south) -- node[right, font=\footnotesize, near end]{$s_t$} (beta.north);
    \draw[arr] (beta) -- node[above, font=\footnotesize]{$\beta_\psi$} (mult);
    \draw[arr] (mod.south) |- node[right, font=\footnotesize, near end]{$I_t^{+,m}$} (mult.north);
    \draw[arr] (mult) -- node[above, font=\footnotesize]{$\alpha\,\beta_\psi\, I_t^{+,m}$} (plus);
    \draw[arr] (buf.east) -| node[right, font=\footnotesize, near start]{$R_t^E$} (plus.north);
    \draw[arr] (plus) -- node[above, font=\footnotesize]{$\bar r_t$} (ppo);

    % Bottom row: Progress loss and feedback to Beta Network
    \node[box, fill=yellow!20, below=1.6cm of mult] (loss) {Progress Loss\\$L_\text{progress}$};
    \draw[darr] (beta.south) |- node[above, font=\footnotesize, near end]{$\beta_\psi(s_t)$} (loss.west);
    \draw[darr] (ppo.south) |- node[right, font=\footnotesize, near end]{$\tau(s_t) \leftarrow |\delta_t|$} (loss.east);
    \draw[darr, blue!60!black] (loss.west) -- ++(-1.6,0) |- node[left, font=\footnotesize, near end]{$\nabla_\psi L_\beta$} (beta.south west);
\end{tikzpicture}
\caption{CARE system architecture. The intrinsic module block stands for any of the three implementations evaluated in this thesis (count-based, ICM, or RIDE); the rest of the pipeline is identical across modules. Solid arrows trace the forward computation that produces the shaped reward $\bar r_t$ used by PPO. Dashed arrows indicate the auxiliary signals used to update the Beta Network $\beta_\psi$ via the learning-progress objective $L_\beta$, with the $\tau(s_t)$ target derived from the magnitude of the critic's TD error $\delta_t$.}
\label{fig:care_arch}
\end{figure}

The shaped reward used throughout this thesis follows directly from Definition~\ref{def:reward_shaping}. For the chosen intrinsic reward module $m$ with rectified intrinsic signal $I_t^{+,m}$ (Section~\ref{sec:icm}), the combined reward optimized by PPO is
\begin{equation}
\bar{r}_t \;=\; R_t^E \;+\; \alpha\, \beta_\psi(s_t)\, I_t^{+,m}.
\label{eq:combined_reward}
\end{equation}
Conceptually, $\beta_\psi(s_t)$ acts as a state-conditional volume control on curiosity: it amplifies intrinsic reward in states where exploration is empirically associated with future extrinsic success and attenuates it elsewhere. The remainder of this chapter formalizes this intuition.

\section{Intrinsic Reward Generation}
\label{sec:intrinsic_modules}

The proposed framework supports three intrinsic reward modules: count-based exploration \cite{bellemare2016count, tang2017exploration}, the Intrinsic Curiosity Module (ICM) \cite{pathak2017curiosity}, and Rewarding Impact-Driven Exploration (RIDE) \cite{raileanu2020ride}. All three differ only in how the raw intrinsic signal $I_t$ is computed at each transition; the rest of the pipeline (standardization, Beta Network, and learning-progress objective) is identical. The empirical evaluation in Chapter~\ref{chap:experiment} reports on all three modules. The remainder of this section describes each module with its raw-reward formulation, followed by the shared standardization step that produces the rectified signal $I_t^+$ fed to the Beta Network.

\subsection{ICM Instantiation}
\label{sec:icm}

\textbf{Forward and Inverse Dynamics Models.} ICM first encodes each state into a latent representation $\phi(s_t) \in \mathbb{R}^d$. It then trains two auxiliary models: a forward dynamics model $f^F_\eta$ that predicts the next latent feature from the current feature and action, and an inverse dynamics model $f^I_\eta$ that predicts the action from two consecutive latent states. The combined ICM loss is
\begin{equation}
L_\text{ICM}(\eta) \;=\;
\alpha_F \, \mathbb{E}\!\left[\tfrac{1}{2}\!\left\| f^F_\eta(\phi_t, a_t) - \phi_{t+1} \right\|_2^2\right]
\;+\; \alpha_I \, \mathbb{E}\!\left[ -\log p_\eta(a_t \mid \phi_t, \phi_{t+1}) \right],
\label{eq:icm_loss}
\end{equation}
with weighting coefficients $\alpha_F, \alpha_I > 0$.

\textbf{Raw Intrinsic Reward.} The raw intrinsic reward at time step $t$ is the forward prediction error
\begin{equation}
I_t \;=\; \tfrac{1}{2}\!\left\| f^F_\eta(\phi_t, a_t) - \phi_{t+1} \right\|_2^2.
\label{eq:icm_raw}
\end{equation}

\subsection{Count-Based Exploration}
\label{sec:count_module}

In the count-based instantiation, the raw intrinsic reward at state $s_t$ is a function of the empirical visit count of that state. Following \cite{tang2017exploration}, we use a random-projection hash $\phi_h : \mathcal{S} \to \mathbb{Z}^{d_h}$ that maps each state to a discrete hash code:
\begin{equation}
\phi_h(s_t) \;=\; \mathrm{round}\!\left( \frac{W^\top s_t}{\Delta} \right),
\label{eq:count_hash}
\end{equation}
where $W \in \mathbb{R}^{d_s \times d_h}$ is a random projection matrix sampled once at initialization, $d_h$ is the hash dimension, and $\Delta > 0$ is the bin width. Let $N(\phi_h(s_t))$ denote the number of times the hash code $\phi_h(s_t)$ has been visited prior to time step $t$. The raw intrinsic reward is the inverse-square-root visitation bonus
\begin{equation}
I_t \;=\; \frac{1}{\sqrt{\max\{N(\phi_h(s_t)),\, 1\}}}.
\label{eq:count_raw}
\end{equation}
This module has no learnable parameters; the hash table $N(\cdot)$ is updated incrementally as new states are visited.

\subsection{Rewarding Impact-Driven Exploration}
\label{sec:ride_module}

In the RIDE instantiation \cite{raileanu2020ride}, the raw intrinsic reward measures the magnitude of representation change between consecutive states in a learned embedding space, divided by an episodic visitation count. Like ICM, RIDE learns an encoder $\phi$ and auxiliary forward and inverse dynamics models $f^F_\eta, f^I_\eta$ trained with the same combined loss $L_\text{ICM}(\eta)$ from Equation~\eqref{eq:icm_loss}; unlike ICM, the forward prediction error is used \emph{only} to train the encoder, never as the intrinsic signal. The raw intrinsic reward is instead the impact of the transition on the latent representation, attenuated by an episodic count:
\begin{equation}
y_t \;=\; \phi(s_t) \in \mathbb{R}^{d_y},
\label{eq:ride_embed}
\end{equation}
\begin{equation}
I_t \;=\; \frac{\bigl\| \phi(s_{t+1}) - \phi(s_t) \bigr\|_2}{\sqrt{\max\{ N_e(\phi_h(s_{t+1})),\, 1 \}}},
\label{eq:ride_raw}
\end{equation}
where $\phi_h$ is the same random-projection hash defined in Equation~\eqref{eq:count_hash} and $N_e(\cdot)$ counts the number of visits to the corresponding hash code \emph{within the current episode} (the count is reset at every episode boundary). The numerator rewards transitions whose representation change is large, capturing the intuition that meaningful actions move the agent through latent space rather than leaving it in place; the denominator suppresses bonuses for revisiting states already encountered in the same episode, which prevents the cyclical wandering that arises when prediction-based bonuses are used in procedurally generated environments. Because the encoder is updated by the same inverse-plus-forward objective as ICM, RIDE inherits ICM's representation learning while replacing the intrinsic signal with an episodic, impact-driven one. We use the same hash dimension $d_h = 32$ and bin width $\Delta = 0.1$ as in the count-based module, and an embedding dimension of $d_y = 256$ matching the ICM encoder.

\subsection{Standardization and Rectification}
\label{sec:standardization}

Regardless of the module, the raw intrinsic reward $I_t$ produced by Equation~\eqref{eq:icm_raw}, \eqref{eq:count_raw}, or \eqref{eq:ride_raw} is non-stationary: its magnitude evolves during training as the underlying generator (forward model, hash counts, or learned encoder) adapts to the agent's experience. To keep the input to the Beta Network bounded in scale and approximately stationary across updates, the same preprocessing step is applied to all three modules.

\begin{definition}[Standardized Intrinsic Reward]
\label{def:standardized_intrinsic}
Given a minibatch with empirical mean $\mu_I$ and standard deviation $\sigma_I$ of the raw intrinsic rewards, and a small constant $\varepsilon > 0$, the \emph{standardized intrinsic reward} is
\begin{equation}
I_t^+ \;=\; \max\!\left( 0,\; \frac{I_t - \mu_I}{\sigma_I + \varepsilon} \right).
\label{eq:icm_standardized}
\end{equation}
\end{definition}

This preprocessing suppresses transitions whose intrinsic signal is below the batch average and retains relatively surprising transitions. We use the generic notation $I_t^{+,m}$ to emphasize that the same pipeline (standardization, Beta Network, progress loss) is applied identically across all three modules; only the upstream raw signal $I_t$ differs.

\section{Correlation-Aware Reward Exploration}
\label{sec:care}

CARE introduces a state-dependent scaling function $\beta_\psi(s_t)$ between the intrinsic reward generator and the policy update. This section formalizes the network, its learning-progress training objective, and the reasons for choosing a first-order optimization scheme over a meta-gradient one.

\subsection{Beta Network}
\label{sec:beta_network}

\begin{definition}[Beta Network]
\label{def:beta_network}
The \emph{Beta Network} is a parameterized map
\begin{equation}
\beta_\psi : \mathcal{S} \to [\beta_\text{min}, \beta_\text{max}], \qquad 0 < \beta_\text{min} < \beta_\text{max} < \infty,
\label{eq:beta_network}
\end{equation}
which assigns a positive, bounded scaling factor to each state. In the implemented system, $\beta_\psi$ is realized as a feed-forward MLP with a 2-layer encoder of width $256$ and a 2-layer head of widths $[256, 128, 1]$ that produces $\log \beta_\psi(s_t)$. The output is mapped to $[\beta_\text{min}, \beta_\text{max}] = [0.1, 2.0]$ via $\beta_\psi(s_t) = \exp\bigl( \mathrm{clamp}(\log \beta_\psi(s_t),\, \log \beta_\text{min},\, \log \beta_\text{max}) \bigr)$. The complete architecture and weight initialization are listed in Appendix~\ref{app:architectures}, Table~\ref{tab:beta_arch}.
\end{definition}

For an intrinsic reward module $m$, the scaled intrinsic reward at time step $t$ is
\begin{equation}
\tilde I_t^{\,m} \;=\; \beta_\psi(s_t)\, I_t^{+,m},
\label{eq:scaled_intrinsic}
\end{equation}
and substituting Equation~\eqref{eq:scaled_intrinsic} into Equation~\eqref{eq:combined_reward} yields the shaped reward used by PPO.

The boundedness assumption built into Definition~\ref{def:beta_network} ensures that the shaped reward cannot grow arbitrarily large relative to the extrinsic signal, regardless of the behavior of the Beta Network during training.

\begin{lemma}[Bounded Shaped Reward]
\label{lem:bounded_shaped}
Let $\beta_\psi$ satisfy Definition~\ref{def:beta_network} and assume the standardized intrinsic reward is bounded almost surely, $|I_t^{+,m}| \le M$ for some constant $M < \infty$. Then for every time step $t$,
\begin{equation*}
\left| \bar{r}_t - R_t^E \right| \;\le\; \alpha\, \beta_\text{max}\, M.
\end{equation*}
Proof in Appendix~\ref{app:proof_bounded}.
\end{lemma}

Lemma~\ref{lem:bounded_shaped} is what allows CARE to be inserted into a PPO training loop without introducing additional instability: the shaped reward differs from the extrinsic reward by a bounded perturbation, so the policy gradient remains bounded under the same regularity conditions that PPO already requires.

\subsection{Learning-Progress Objective}
\label{sec:progress_loss}

The Beta Network is trained so that $\beta_\psi(s_t)$ becomes large at states where the value function exhibits large temporal-difference (TD) error and small at states where the value function is well-fitted. The intuition is that a large $|\delta_t|$ marks a state where extrinsic learning is still in progress, so additional intrinsic exploration there is informative; conversely, when $|\delta_t|$ is small, the agent already has a stable value estimate and additional curiosity bonus would not improve learning.

For each transition in the rollout buffer, the one-step TD error of the extrinsic value function is
\begin{equation}
\delta_t \;=\; R_t^E \;+\; \gamma\, V_\mu(s_{t+1}) (1 - d_t) \;-\; V_\mu(s_t),
\label{eq:td_error}
\end{equation}
where $d_t \in \{0, 1\}$ is the episode termination flag and $V_\mu$ is the current critic. The per-state target $\tau(s_t)$ is then defined as a piecewise normalization of $|\delta_t|$ onto the Beta Network's output range:
\begin{equation}
\tau(s_t) \;=\; \begin{cases}
\beta_\text{min} + (\beta_\text{max} - \beta_\text{min}) \dfrac{|\delta_t|}{\max_{t'} |\delta_{t'}| + \varepsilon}, & \text{if } \max_{t'} R^E_{t'} > 0, \\[6pt]
\beta_0, & \text{otherwise (cold-start fallback).}
\end{cases}
\label{eq:beta_target}
\end{equation}
The cold-start fallback addresses the regime in which a batch contains no positive extrinsic reward; in that case $|\delta_t|$ reflects mainly initialization noise of the critic, and would be a poor target for $\beta_\psi$. Defaulting $\tau$ to the reference value $\beta_0 = 1$ in that regime makes CARE behave like a fixed-coefficient method until extrinsic signal becomes informative.

The learning-progress loss is the mean-squared deviation of $\beta_\psi$ from this target:
\begin{equation}
L_\text{progress}(\psi) \;=\; \mathbb{E}\!\left[ \bigl( \beta_\psi(s_t) - \tau(s_t) \bigr)^2 \right].
\label{eq:progress_loss}
\end{equation}
A log-space regularizer pulls $\beta_\psi$ toward the reference value $\beta_0 = 1$ to prevent degenerate scaling and to control variance during early training:
\begin{equation}
L_\text{reg}(\psi) \;=\; \mathbb{E}\!\left[ \bigl( \log \beta_\psi(s_t) - \log \beta_0 \bigr)^2 \right].
\label{eq:reg_loss}
\end{equation}
The complete Beta-network objective is
\begin{equation}
L_\beta(\psi) \;=\; \gamma_p\, L_\text{progress}(\psi) \;+\; \lambda_\text{reg}\, L_\text{reg}(\psi),
\label{eq:beta_loss}
\end{equation}
with $\gamma_p > 0$ the progress weight and $\lambda_\text{reg} > 0$ the regularization weight.

The progress loss has a clean statistical interpretation, and its gradient has a clean directional one.

\begin{proposition}[Target Recovery]
\label{prop:target_recovery}
Assume the Beta Network has sufficient expressive capacity to represent any measurable function $\mathcal{S} \to [\beta_\text{min}, \beta_\text{max}]$. In the population limit and with $\lambda_\text{reg} = 0$, the unique minimizer $\psi^\star$ of $L_\beta(\psi)$ satisfies
\begin{equation*}
\beta_{\psi^\star}(s) \;=\; \mathbb{E}\!\left[\, \tau(s_t) \,\big|\, s_t = s \right]
\end{equation*}
for every state $s$ visited with positive probability.
Proof in Appendix~\ref{app:proof_target}.
\end{proposition}

Proposition~\ref{prop:target_recovery} states that, in the population limit, the optimal Beta Network simply outputs the conditional mean of the target $\tau$ at each state. Because $\tau$ is a monotone normalization of $|\delta_t|$, this means that $\beta_{\psi^\star}(s)$ is large precisely at states where the extrinsic value function has, on average, larger TD error. The next result describes the gradient direction during training.

\begin{proposition}[Gradient Direction]
\label{prop:gradient_direction}
For every state $s \in \mathcal{S}$ visited with positive probability,
\begin{equation*}
\frac{\partial L_\text{progress}}{\partial \beta_\psi(s)}
\;=\;
2\, \mathbb{P}(s_t = s)\, \Bigl( \beta_\psi(s) \;-\; \mathbb{E}[\tau(s_t) \mid s_t = s] \Bigr).
\end{equation*}
In particular, gradient descent on $L_\text{progress}$ moves $\beta_\psi(s)$ monotonically toward the conditional mean target $\mathbb{E}[\tau(s_t) \mid s_t = s]$.
Proof in Appendix~\ref{app:proof_gradient}.
\end{proposition}

Proposition~\ref{prop:gradient_direction} makes the central design intuition of CARE explicit: a gradient step adjusts $\beta_\psi(s)$ toward the local average of $\tau$, which by Equation~\eqref{eq:beta_target} is a normalized measure of how unsettled the value function still is at state $s$. The Beta Network is therefore supervised by the agent's own learning progress as observed through the critic's TD error.

\subsection{First-Order Optimization vs. Meta-Gradient}
\label{sec:first_order}

A natural alternative to the learning-progress objective is to learn the intrinsic reward weighting via meta-gradients, as in Learning Intrinsic Rewards for Policy Gradient methods (LIRPG) \cite{zheng2018lirpg}. In that approach, the parameters of the intrinsic reward function are updated so that, after a policy update using the shaped reward, the resulting policy maximizes the extrinsic objective. This requires differentiating through one or more inner policy updates, which is computationally expensive and can be unstable in deep reinforcement learning settings.

CARE replaces the meta-gradient by a first-order objective that supervises $\beta_\psi$ directly through the target $\tau(s_t)$ derived from the critic's TD error. The cost difference is quantified by the following result.

\begin{proposition}[Computational Cost]
\label{prop:cost}
Per Beta-network update of size $T$, let $K$ denote the number of inner policy update steps used in a meta-gradient estimator, $|\theta|$ the number of policy parameters, and $|\psi|$ the number of Beta-network parameters. Then a meta-gradient training step has worst-case time complexity $O(K\, T\, |\theta|)$, while a CARE first-order training step has time complexity $O(T\, |\psi|)$. In particular, when $|\psi| \ll |\theta|$ and $K \ge 1$, the CARE update is asymptotically cheaper by a factor of $K\, |\theta| / |\psi|$.
Proof in Appendix~\ref{app:proof_cost}.
\end{proposition}

In the implemented system, $|\psi|$ is roughly two orders of magnitude smaller than $|\theta|$, so the cost reduction is substantial. The price of this efficiency is a weaker form of supervision: instead of optimizing $\beta_\psi$ directly for downstream extrinsic performance, CARE optimizes it for a TD-error-based target that serves as a proxy for that performance. The next result describes two regimes in which this proxy reduces to the trivial fixed-coefficient case.

\begin{corollary}[Degeneration to Fixed Coefficient]
\label{cor:degeneration}
Consider the objective $L_\beta(\psi)$ in Equation~\eqref{eq:beta_loss}. There are two regimes in which every minimizer $\psi^\star$ satisfies $\beta_{\psi^\star}(s) \to \beta_0 = 1$ uniformly in $s$, and the shaped reward in Equation~\eqref{eq:combined_reward} reduces to the fixed-coefficient form $\bar r_t = R_t^E + \alpha\, I_t^{+,m}$:
\begin{enumerate}
    \item[(i)] \textbf{Cold-start regime}: when $\max_{t} R_t^E \le 0$ within the rollout buffer, the target $\tau$ in Equation~\eqref{eq:beta_target} reduces to $\beta_0$ by definition, and the unique minimizer of $L_\text{progress}$ is $\beta_\psi \equiv \beta_0$.
    \item[(ii)] \textbf{Strong regularization regime}: as $\lambda_\text{reg} \to \infty$, the regularization term dominates and forces $\log \beta_\psi(s) \to 0$ in $L^2$ under the state-visitation distribution, equivalently $\beta_\psi(s) \to \beta_0 = 1$.
\end{enumerate}
Proof in Appendix~\ref{app:proof_degeneration}.
\end{corollary}

Corollary~\ref{cor:degeneration} has both a theoretical and a practical reading. Theoretically, it shows that CARE strictly contains fixed-coefficient PPO+ICM as a limiting case, so the framework cannot perform worse than that baseline once the regularization is appropriately tuned and the cold-start fallback is in place. Practically, the cold-start branch is essential in extremely sparse environments: when the agent has not yet encountered any extrinsic reward, the TD error reflects only initialization noise of the critic, and falling back to $\beta_0 = 1$ prevents CARE from amplifying that noise into the policy update.

\section{Algorithm}
\label{sec:algorithm}

This section formalizes the data structures, loss functions, and training procedure used to optimize the system end to end.

\subsection{Rollout Buffer}
\label{sec:rollout_buffer}

Each PPO update operates on an on-policy rollout buffer of length $T$ collected by the current policy. A single transition is the tuple
\begin{equation}
\xi_t \;=\; \bigl( s_t,\; a_t,\; \log \pi_\theta(a_t \mid s_t),\; R_t^E,\; s_{t+1},\; d_t \bigr),
\label{eq:transition_tuple}
\end{equation}
where $d_t \in \{0, 1\}$ is the episode termination flag. The buffer of length $T$ is stored as the tensors
\begin{equation*}
\mathcal{S}_\text{tensor} \in \mathbb{R}^{T \times d_s}, \quad
\mathcal{A}_\text{tensor} \in \mathbb{N}^{T}, \quad
\mathcal{R}_\text{tensor} \in \mathbb{R}^{T}, \quad
\mathcal{S}'_\text{tensor} \in \mathbb{R}^{T \times d_s}, \quad
\mathcal{D}_\text{tensor} \in \{0,1\}^{T},
\end{equation*}
together with the log-probabilities $\mathcal{P}_\text{tensor} \in \mathbb{R}^{T}$. Following the implementation defaults, $T = 4 \times \text{max\_ep\_len}$ steps are collected before each PPO update.

\subsection{Loss Functions}
\label{sec:loss_functions}

Four distinct losses are minimized at each update cycle. They share the same rollout buffer but are optimized with respect to disjoint parameter groups, so no joint optimization is required.

\textbf{PPO Actor Loss $L^\text{PPO}(\theta)$.} Define the importance ratio $\rho_t(\theta) = \pi_\theta(a_t \mid s_t) / \pi_{\theta_\text{old}}(a_t \mid s_t)$ and let $\hat A_t$ denote the advantage estimate computed by Generalized Advantage Estimation (GAE) \cite{schulman2016gae} from the shaped reward $\bar r_t$. The clipped surrogate objective \cite{schulman2017ppo} is
\begin{equation}
L^\text{PPO}(\theta) \;=\;
-\,\mathbb{E}_t\!\left[
\min\!\Bigl(
\rho_t(\theta)\, \hat A_t,\,
\text{clip}\bigl(\rho_t(\theta),\, 1-\epsilon,\, 1+\epsilon\bigr)\, \hat A_t
\Bigr)\right].
\label{eq:ppo_loss}
\end{equation}

\textbf{Critic Loss $L^V(\mu)$.} The value network $V_\mu$ is fitted by mean-squared error against the GAE return target $\hat V_t$:
\begin{equation}
L^V(\mu) \;=\; \mathbb{E}_t\!\left[ \bigl( V_\mu(s_t) - \hat V_t \bigr)^2 \right].
\label{eq:critic_loss}
\end{equation}

\textbf{ICM Loss $L_\text{ICM}(\eta)$.} As specified in Equation~\eqref{eq:icm_loss}.

\textbf{Beta Loss $L_\beta(\psi)$.} As specified in Equation~\eqref{eq:beta_loss}.

The four losses correspond to four parameter groups: $\theta$ (actor), $\mu$ (critic), $\eta$ (ICM), and $\psi$ (Beta Network). Gradients flow through each loss with respect to its own parameter group only; in particular, the Beta Network is not updated through $L^\text{PPO}$ and the policy is not updated through $L_\beta$.

\subsection{Training Procedure}
\label{sec:training_procedure}

Algorithm~\ref{alg:care} specifies the complete CARE training loop. One iteration of the outer loop corresponds to a single PPO update cycle.

\begin{algorithm}[H]
\caption{CARE training loop with PPO and an intrinsic reward module.}
\label{alg:care}
\begin{algorithmic}[1]
\State Initialize policy $\theta$, critic $\mu$, intrinsic-module parameters $\eta$ (if any), Beta Network $\psi$
\For{$\text{update} = 1, 2, \dots, U$}
    \State Collect $T$ on-policy transitions $\{\xi_t\}_{t=0}^{T-1}$ using $\pi_\theta$
    \State Compute raw intrinsic rewards $I_t$ via the module-specific equation in Section~\ref{sec:intrinsic_modules}
    \State Standardize and rectify to obtain $I_t^+$ via Equation~\eqref{eq:icm_standardized}
    \State Compute extrinsic TD errors $\delta_t$ via Equation~\eqref{eq:td_error} and targets $\tau(s_t)$ via Equation~\eqref{eq:beta_target}
    \State Update intrinsic-module parameters $\eta$ (no-op if module is non-parametric, e.g., count-based)
    \State Update Beta Network: $\psi \gets \psi - \psi_\text{lr}\, \nabla_\psi L_\beta(\psi)$
    \State Compute shaped rewards $\bar r_t = R_t^E + \alpha\, \beta_\psi(s_t)\, I_t^+$
    \State Compute GAE advantages $\hat A_t$ from $\{\bar r_t\}$ and $V_\mu$
    \For{$\text{epoch} = 1, 2, \dots, K$}
        \State Update actor: $\theta \gets \theta - \theta_\text{lr}\, \nabla_\theta L^\text{PPO}(\theta)$
        \State Update critic: $\mu \gets \mu - \mu_\text{lr}\, \nabla_\mu L^V(\mu)$
    \EndFor
\EndFor
\end{algorithmic}
\end{algorithm}

The algorithm preserves the standard PPO update structure (lines 11--14) and inserts only two additional optimization steps (lines 7--8) before the shaped reward is computed. This modularity is important: the same template applies if ICM is replaced by a count-based bonus or by RIDE, with only Equations~\eqref{eq:icm_raw}--\eqref{eq:icm_standardized} being substituted by the corresponding module-specific computation. Concrete numerical values for all hyperparameters and network architectures used in the experiments are listed in Appendix~\ref{app:implementation}.
