\chapter{Experimental Setup and Results}
\label{chap:experiment}

This chapter describes the experimental setup used to evaluate the proposed methodology. It first presents the benchmark environments and the computing infrastructure, then specifies the experimental configuration, the baselines, and the performance evaluation metrics. The empirical evaluation across all three intrinsic reward families (count-based, ICM, and RIDE) is the subject of Section~\ref{sec:empirical_results}. Settings reported here are the canonical configuration shared by every reported run; the underlying hyperparameter values are tabulated in Appendix~\ref{app:implementation}.

\section{Benchmark Environments}
\label{sec:benchmark_environments}

The empirical evaluation is conducted on six sparse-reward MiniGrid environments \cite{chevalier2018minigrid}. These environments are chosen because they cover qualitatively different forms of exploration difficulty while sharing a common gridworld interface, which keeps the comparison across intrinsic reward modules controlled.

\begin{table}[H]
\centering
\caption{MiniGrid environments used in the experiments.}
\label{tab:envs}
\begin{tabular}{|m{4cm}|m{9cm}|}
\hline
\textbf{Environment} & \textbf{Purpose in the evaluation} \\ \hline
DoorKey-8x8 & Tests whether the agent can discover a key, unlock a door, and shift from exploration to goal-directed behavior. \\ \hline
Empty-16x16 & Represents an extreme sparse-reward setting with almost no intermediate feedback before the goal is reached. \\ \hline
KeyCorridor-S3R3 & A multi-room layout with several keys and locked doors; assesses sustained exploration over a longer horizon. \\ \hline
LavaCrossing-S9N3 & Requires safe navigation around lava; penalizes irrelevant exploration that leads to early termination. \\ \hline
RedBlueDoors-8x8 & Requires opening two doors in the correct order, testing whether the method can suppress irrelevant exploration. \\ \hline
UnlockPickup & Extends DoorKey with an additional object manipulation stage, creating a hierarchical sparse-reward task. \\ \hline
\end{tabular}
\end{table}

These tasks share several useful properties for studying adaptive intrinsic reward scaling: sparse terminal rewards, egocentric partial observations, step penalties that discourage random motion, and stochastic initialization across random seeds. Together, they form a test bed in which different intrinsic reward modules can be compared under the same policy optimization backbone and the same observation interface.

\section{Computing Environment}
\label{sec:computing_environment}

All experiments were carried out on a single machine with the following configuration:
\begin{itemize}
    \item \textbf{Hardware}: \texttt{[CPU model, e.g., Intel Core i7-13700K, 16 cores]}; \texttt{[GPU model, e.g., NVIDIA RTX 4070, 12 GB VRAM]}; \texttt{[RAM, e.g., 32 GB]}.
    \item \textbf{Operating system}: Windows 11.
    \item \textbf{Software stack}: Python 3.12, PyTorch with CUDA support, \texttt{gymnasium} with the \texttt{minigrid} extension, \texttt{numpy}, \texttt{pandas}, and \texttt{matplotlib}.
    \item \textbf{Wall-clock budget}: each seeded run consists of $10^6$ environment steps; the approximate runtime per seed is \texttt{[\,\!\!fill in after benchmarking,\!\!]} hours on the configuration above.
\end{itemize}

The training scripts automatically dispatch the policy and auxiliary networks to GPU when CUDA is available and fall back to CPU otherwise. All experiments use deterministic random seeds for reproducibility, and all training logs and checkpoint files are saved to disk every $10^5$ steps.

\section{Experimental Configuration}
\label{sec:experimental_config}

All reported runs share the same PPO backbone and the same intrinsic-reward processing pipeline so that the comparison across modules and across scaling strategies is fair. The complete set of hyperparameter values is tabulated in Appendix~\ref{app:hyperparameters}; the most relevant high-level settings are summarized here.

No separate global intrinsic strength coefficient is used: the Beta Network output $\beta_\psi(s_t)$ alone determines the intrinsic-reward scale, and in fixed-$\beta$ conditions the same scalar $\beta$ plays this role. In the CARE conditions, the Beta Network is realized as a two-layer encoder of dimension $256$ followed by a $128$-unit head producing $\log \beta_\psi$, which is clipped so that $\beta_\psi(s_t) \in [\beta_\text{min}, \beta_\text{max}] = [0.001, 0.1]$. This range is chosen to span an order of magnitude either side of the optimal fixed coefficients reported for sparse MiniGrid tasks (approximately $\beta \approx 0.005$ for count-based bonuses and $\beta \approx 0.05$ for ICM \cite{andres2025impact}), so that the adaptive variant has the freedom to recover, undershoot, or overshoot any of those values. The learning-progress objective in Equation~\eqref{eq:beta_loss} uses regularization weight $\lambda_\text{reg} = 10^{-3}$, reference value $\beta_0 = 0.01$ (the geometric centre of the interval), gradient clipping at $L_2$ norm $1.0$, and $L_2$ weight decay of $10^{-6}$ on the Beta Network parameters. The Beta Network is updated with Adam at learning rate $5 \times 10^{-4}$. In the fixed-$\beta$ conditions, the Beta Network is replaced by a constant scalar $\beta \in \{0.001, 0.005, 0.05, 0.1\}$ and the learning-progress update is disabled; the rest of the pipeline (intrinsic generation, standardization, and the shaped reward formula in Equation~\eqref{eq:combined_reward}) is unchanged.

The base PPO algorithm uses $K = 80$ epochs per update, clip parameter $\epsilon = 0.2$, discount factor $\gamma = 0.99$, GAE bias-variance trade-off $\lambda = 0.95$, actor learning rate $3 \times 10^{-4}$, and critic learning rate $10^{-3}$. The rollout buffer length is $T = 4{,}000$ steps, and each seed is trained for $10^6$ environment steps. Each configuration is repeated over five independent random seeds $\{1, 2, 3, 4, 5\}$.

\section{Baselines}
\label{sec:baselines}

The experimental framing of this thesis treats CARE as a module-agnostic adaptive layer that can be paired with any of the three representative intrinsic reward families: count-based bonuses, the Intrinsic Curiosity Module (ICM), and Rewarding Impact-Driven Exploration (RIDE). To isolate the contribution of CARE from both the contribution of the underlying intrinsic signal and the choice of any particular fixed scaling coefficient, the baselines are organized into a $1 + 12 + 3$ matrix as summarized in Table~\ref{tab:baselines}: one no-intrinsic control, twelve fixed-$\beta$ configurations spanning a two-order-of-magnitude sweep across the three modules, and three CARE variants.

\begin{table}[H]
\centering
\caption{Baseline conditions evaluated on every environment.}
\label{tab:baselines}
\begin{tabular}{|m{1.8cm}|m{4.4cm}|m{7.2cm}|}
\hline
\textbf{Label} & \textbf{Configuration} & \textbf{Purpose in the comparison} \\ \hline
$B_0$ & PPO (no intrinsic reward) & Lower bound on exploration ability without any curiosity signal. \\ \hline
$F_{m,\beta}$ & PPO + module $m$ with constant $\beta$, $m \in \{\text{Count}, \text{ICM}, \text{RIDE}\}$, $\beta \in \{0.001,\, 0.005,\, 0.05,\, 0.1\}$ & Fixed-coefficient sweep for each module. The four values span two orders of magnitude and bracket the optimal coefficients reported for sparse MiniGrid tasks ($\beta \approx 0.005$ for count-based and $\beta \approx 0.05$ for ICM \cite{andres2025impact}). \\ \hline
$C_\text{Count}$ & PPO + count-based + CARE & Count-based bonus with learned $\beta_\psi(s_t)$ from the Beta Network. \\ \hline
$C_\text{ICM}$ & PPO + ICM + CARE & ICM bonus with learned $\beta_\psi(s_t)$. \\ \hline
$C_\text{RIDE}$ & PPO + RIDE + CARE & RIDE bonus with learned $\beta_\psi(s_t)$. \\ \hline
\end{tabular}
\end{table}

This design isolates three distinct effects:
\begin{itemize}
    \item \textbf{$B_0$ vs $F_{m,\beta}$} measures whether intrinsic motivation in module $m$ at coefficient $\beta$ helps or hurts on top of the extrinsic-only PPO baseline. Sweeping $\beta$ over two orders of magnitude around the literature-reported optima exposes how sensitive each module is to manual tuning within the relevant regime.
    \item \textbf{$F_{m,\beta}$ vs $C_m$} measures the contribution of CARE's adaptive scaling, holding the underlying intrinsic signal fixed. The cleanest test is the comparison of $C_m$ against the \emph{best} fixed-$\beta$ value for module $m$ on each environment: if CARE matches or exceeds the best post-hoc fixed coefficient \emph{without} that coefficient having to be searched, the case for state-dependent scaling is established.
    \item \textbf{$C_\text{Count}$ vs $C_\text{ICM}$ vs $C_\text{RIDE}$} reveals how the same CARE layer interacts with structurally different intrinsic generators (count-based visitation, ICM forward error, and RIDE impact-with-episodic-counts), addressing the research question of module-agnosticism stated in Chapter~1.
\end{itemize}

With $1 + 12 + 3 = 16$ conditions per environment and five seeds per condition, the full evaluation comprises $16 \times 5 = 80$ runs per environment and $80 \times 6 = 480$ runs across the six environments.

\section{Performance Evaluation Metrics}
\label{sec:metrics}

Four metrics are used to quantify and compare the conditions defined above.

\textbf{Episode return.} Let $R^E_t$ be the extrinsic reward at time $t$ in evaluation episode $e$. The episode return for that episode is $\mathcal{R}_e = \sum_t R^E_t$. The reported curves are the mean of $\mathcal{R}_e$ over a sliding window of evaluation episodes around each checkpoint, averaged across the five seeds. Episode return is the primary metric of task success.

\textbf{Sample efficiency.} For each environment, a target return $r^\star$ is fixed at a fraction (typically $0.7$) of the asymptotic return achieved by the strongest baseline. The sample efficiency of a method is then the smallest number of environment steps $n^\star$ at which its mean evaluation return first reaches $r^\star$:
\begin{equation*}
n^\star \;=\; \min\bigl\{ n : \overline{\mathcal{R}}(n) \ge r^\star \bigr\},
\end{equation*}
with $n^\star = +\infty$ if the threshold is never reached within the $10^6$-step budget. Lower values of $n^\star$ indicate faster learning.

\textbf{Learning stability.} For each method and each timestep $n$, the learning stability is reported as the across-seed standard deviation $\sigma(n)$ of the mean evaluation return at that timestep:
\begin{equation*}
\sigma(n) \;=\; \sqrt{\frac{1}{|S| - 1} \sum_{s \in S} \bigl( \overline{\mathcal{R}}_s(n) - \overline{\mathcal{R}}(n) \bigr)^2},
\end{equation*}
where $S$ is the set of seeds. Lower $\sigma(n)$ indicates more reproducible behavior.

\textbf{Learned scaling behavior.} The learned function $\beta_\psi$ is analyzed both globally and locally. Globally, the empirical distribution of $\beta_\psi(s_t)$ over the rollout buffer is summarized by its mean and selected percentiles ($25\%, 50\%, 75\%$) at fixed checkpoints throughout training. Locally, per-environment heatmaps of $\beta_\psi(s)$ projected onto the agent's two-dimensional position are produced from the trajectory data saved every $10^5$ steps, providing a qualitative view of where in state space CARE allocates more exploration influence.

\section{Empirical Results}
\label{sec:empirical_results}

\textit{Empirical results across the six benchmark environments and the sixteen conditions defined in Section~\ref{sec:baselines} will be reported here once the experimental runs are complete. The remainder of this section is organized into the following subsections, which will be populated with reward curves, summary tables, and qualitative analyses of $\beta_\psi(s)$.}

\subsection{Module-Level Results: Fixed-$\beta$ Sweep vs CARE}
\textit{Per-module reward curves and summary statistics. For each module $m \in \{\text{Count}, \text{ICM}, \text{RIDE}\}$, the five $F_{m,\beta}$ curves and the $C_m$ curve are plotted on a single panel per environment, with $B_0$ as a shared reference. Tables report the best fixed-$\beta$ value per environment and the gap to $C_m$.}

\subsection{Effect of CARE Across Modules}
\textit{Cross-module comparison summarizing where CARE provides the largest gain over the best fixed coefficient, where it is neutral, and where the cold-start fallback dominates. Includes a heatmap of CARE-vs-best-fixed gains across the six environments and three modules.}

\subsection{Sensitivity to Fixed $\beta$}
\textit{Analysis of how sharply performance varies across the swept range $\beta \in \{0.001,\, 0.005,\, 0.05,\, 0.1\}$ for each module, motivating the need for adaptive scaling: modules whose performance varies strongly with $\beta$ are the ones where CARE is expected to help most.}

\subsection{Behavior of the Learned Scaling Function}
\textit{Distributional and trajectory-level analysis of $\beta_\psi(s_t)$, including evolution over training and qualitative state heatmaps.}

\subsection{Discussion}
\textit{Synthesis of the empirical findings in light of the formal results in Chapter~3 and the literature reviewed in Chapter~2.}
